Functionally Important Residues from Graph Analysis of Coevolved Dynamic couplings

Version 1

Abstract

Abstract

The relationship between protein dynamics and function is essential for understanding biological processes and developing effective therapeutics. Functional sites within proteins are critical for activities such as substrate binding, catalysis, and structural changes. Existing computational methods for the predictions of functional residues are trained on sequence, structural and experimental data, but they do not explicitly model the influence of evolution on protein dynamics. This overlooked contribution is essential as it is known that evolution can fine tune protein dynamics through compensatory mutations, either to improve the proteins’ performance or diversify its function while maintaining the same structural scaffold. To model this critical contribution, we introduce DyNoPy, a computational method that combines residue coevolution analysis with molecular dynamics (MD) simulations, revealing hidden correlations between functional sites. DyNoPy constructs a graph model of residue-residue interactions, identifies communities of key residue groups and annotates critical sites based on their roles. By leveraging the concept of coevolved dynamical couplings—residue pairs with critical dynamical interactions that have been preserved during evolution—DyNoPy offers a powerful method for predicting and analysing protein evolution and dynamics. We demonstrate the effectiveness of DyNoPy on SHV-1 and PDC-3, chromosomally encoded β-lactamases linked to antibiotic resistance, highlighting its potential to inform drug design and address pressing healthcare challenges.

Introduction

Quantifying the contribution of individual residues or residue groups to protein function is important to estimate the pathogenic effect of mutations (1). Identifying the functional roles of individual residues has primarily been done through mutagenesis experiments (2). Bioinformatics methods have complemented these approaches through analysis of multiple sequence alignments (MSA) of homologous proteins and structural data (3–8). Among these methods, computational techniques that can decode inter-residue evolutionary relationships from MSAs have paved the way for machine learning (ML) based strategies that can predict protein structure (9–12), stability (13), and function (7) and extend the scope of computational protein design (14–16). A most recent approach has combined experimental data from three proteins, NUDT15, PTEN and CYP2C9, on stability and function with sequence and structural features to train a ML model to predict functional sites (17).

Functional sites are often regulated by both, local and global interactions. Changes in these interactions are instrumental for functional events like substrate binding, catalysis, and conformational changes (18). While techniques described earlier have exploited the sequence and structure relationship of proteins, they do not directly consider the time evolution of functional events in their predictions. The development of physical models of protein dynamics and the increase in available computational power has stimulated the adoption of computational techniques (19, 20) to investigate the conformational dynamics of proteins, an essential component of the many biological functions (21, 22). It has also been established that evolution through compensatory mutations in dynamic regions, like hinges and loops, can fine tune protein structural dynamics and introduce promiscuity, thereby diversifying biological function. Assuming that protein functional dynamics is conserved during evolution, significant information on dynamic regions and substrate recognition sites should be recoverable using inter residue coevolution scores extracted from MSAs (23, 24). Coevolution analysis and Molecular Dynamics (MD) simulations have independently (25) and synergistically been combined in the past to identify important residues for function (26–30). Yet a method that combines hidden information on dynamics from evolution with direct information on local and global dynamics from conformational ensembles from MD is not available.

Here, we present DyNoPy, a computational method that can extract hidden information on functional sites from the combination of pairwise residue coevolution data and powerful descriptors of dynamics extracted from the analysis of MD ensembles. The method can detect coevolved dynamic couplings, i.e. residue pairs with critical dynamical interactions that have been preserved during evolution. These pairs are extracted from a graph model of residue-residue interactions. Communities of important residue groups are detected, and critical sites are identified by their eigenvector centrality in the graph (Fig. 1). We demonstrate the power of this approach on SHV-1 and PDC-3 β-lactamases of major clinical importance (31, 32). DyNoPy successfully detects residue couplings that align with previous studies, guide in the explanations of mutation sites with previously unexplained mechanisms and provide predictions on plausible important sites for the emergence of clinically relevant variants.

Overview of the DyNoPy workflow.

Results and Discussion

β-lactamases are a group of enzymes capable of hydrolysing β-lactams, conferring resistance to β-lactam antibiotics (33). These enzymes are evolving rapidly, as single amino acid substitutions are sufficient to drive their evolution and increase their catalytic spectrum and inhibitor resistance profile (34). The widespread dissemination of β-lactamases across different bacterial species and their extensive emergence highlight their global impact on antibiotic resistance (35). The rapid evolution of β-lactamases and their clinical significance (34) makes them an ideal target for evaluating the robustness of DyNoPy.

In this study, we applied DyNoPy to two model enzymes from different β-lactamase families: class A β-lactamase SHV-1 (a chromosomally encoded enzyme in Klebsiella pneumoniae) and class C β-lactamase PDC-3 (a chromosomally encoded enzyme in Pseudomonas aeruginosa) (31, 32) (Fig. S1 and S2). Both class A and class C β-lactamases comprise an α/β domain and an α helical domain, with the active site situated in between (36, 37). Moreover, both enzymes target the carbonyl carbon of the β-lactams using a highly conserved serine residue (38, 39). Despite these similarities, the structures of class A and class C β-lactamases are remarkably different (Fig. 2).

Structural Comparison of SHV-1 (PDB ID: 3N4I) and PDC-3 (PDB ID: 4HEF) β-Lactamases.

Catalytic serine S70 (SHV-1) and S64 (PDC3) are highlighted using stick representation. Important loops surrounding the active site are highlighted in red. In SHV-1, highlighted loops are the α3-α4 loop (residues 101-111), the Ω-loop (residues 164-179), and the hinge region (residues 213-218). In PDC-3, highlighted loops are the Ω-loop (residues 183-226) and the R2-loop (residues 280-310).

SHV-1 is a very well characterised enzyme with wealth of information on mutations and their corresponding effects on protein function. In contrast, the information available on PDC-3 remains limited. Detailed structural information on these enzymes can be found in the supporting materials. Essential catalytic residues in SHV-1 are: S70, K73, S130, E166, N170, K234, G236, and A237 (40) and conserved catalytic residues in PDC-3 include S64, K67, Y150, N152, K315, T316, and G317. Highly conserved stretches of 3-9 hydrophobic residues, annotated as hydrophobic nodes, exists in class A β-lactamases and have been proven to be essential for protein stability (41). Residues defined as belonging to hydrophobic nodes within SHV-1 are listed in Supporting Table S1.

In SHV-1, the predominant extended spectrum β-lactamase (ESBL) substitutions occur at L35, G238, and E240, while R43, E64, D104, A146, G156, D179, R202, and R205 appear in ESBLs with lower frequency (42). Mutations at M69, S130, A187, T235, and R244 are known to induce inhibitor resistance in the enzyme (43). In PDC-3, substitutions primarily occur on the Ω-loop, enhancing its flexibility to accommodate the bulky side chains of antibiotics, while deletions are more common in the R2-loop (39). The predominant Ω-loop mutations isolated from clinics are found at positions V211, G214, E219, and Y221 (44).

Emergence of highly conserved dynamic couplings

DyNoPy builds a pairwise model of conserved dynamic couplings detected by combining coevolution scores and information on functional motions into a score Jij (see Methods and Fig. 1). To this end a dynamics descriptor should be selected. When the descriptor is associated with functional conformational changes, it is expected that functionally relevant couplings will report higher scores. Dynamics descriptors can be selected from commonly used geometrical collective variables (CVs) for the analysis of MD trajectories (see Methods). As expected, the average J matrix score varies across the different CVs, with some of them showing no signal of dynamic coupling (Supporting Fig. S4C).

SHV-1 and PDC-3 exhibit distinct dynamics, requiring a different choice of the CV that best captures the functional dynamics. For SHV-1, the global first principal component (PC1) proved to be the most effective feature, identifying 571 residue pairs with a Jij value greater than 0. Conversely, PDC-3 requires selection of more localized features that can extract the Ω-loop dynamics from the overall protein motion. Among the dynamic descriptors, the partial first time-lagged component (TC1_partial) performed best for PDC-3, detecting 216 residue pairs with a Jij value greater than 0. Consequently, PC1 and TC1_partial were selected to build the J matrix for SHV-1 and PDC-3, respectively. The performance of all 12 CVs for each protein was assessed and listed in the Supporting Table S2.

The importance of dynamical information is evident when coevolution couplings (γij) and conserved dynamic couplings (Jij) are compared: the number of non-zero couplings decrease from 40% to <2% of total residue pairs in the protein (Fig. S4D) when information from the dynamics descriptor is added. Thus, the inclusion of protein dynamics in coevolution studies acts as an effective filter that rules out residue pairs that do not have significant correlations with functional motions. Moreover, when relying only on γij, all the residues in SHV-1 and PDC-3 are included within four identified communities (Supporting Table S3), suggesting that coevolution scores (γij) alone do not effectively discriminate residues relevant for protein functions. Furthermore, it would be hard to distinguish critical core residues for each community using only γij, as the eigenvector centrality (EVC) values for the residues do not show remarkable differences (Supporting Fig. S5A and S5B). This means that detailed dynamic investigation of the top residues is needed to determine which pairs should be picked up and further analysed. On the other hand, it is much easier to identify essential residues based on J scores calculated, as clear outliers with significantly higher EVC values could be seen for almost all communities (Supporting Fig. S5C and S5D) (25, 45). In conclusion, the lack of specificity in the statistically based coevolution analysis supports the choice of incorporating a score for the correlation between residue interactions and dynamic behaviours that enables deconvolution of community information.

DyNoPy reveals critical residues and predicts evolutionary pathways in SHV-1

DyNoPy identified eight significant communities of strongly coupled residues within SHV-1 (Supporting Fig. S4A) with all crucial catalytic residues and critical substitution sites previously mentioned participating in one of these communities with the exceptions of R43, R202, and S130.

Residues previously known to have critical role in function or conferring ESBLs/IRBLs phenotype are either directly coupled to protein dynamics or act as a central hub. The hubs interact with residues with either a role in catalysis or structural stability through their membership of hydrophobic nodes (31). Furthermore, DyNoPy identified key positions (L162 and N136) within some communities that are known to undergo substitutions, conferring an ESBL phenotype in other class A β-lactamases. These substitutions have not yet emerged in the SHV family, providing insightful predictions about the potential future evolution of the enzyme. Detailed description of the other three communities is provided in the supporting information (Fig. S6).

DyNoPy predicts mutation hotspots in SHV-1

DyNoPy detects critical mutation sites (L162 and N136) that are known to extend the range of substrates in other class A β-lactamases but have not yet emerged as variants in the SHV family. These sites have not been modified in SHV family because of their plausible central role within the communities as they are mediating couplings with key functional residues essential for catalytic activity and structural stability, indicating their critical role in protein function and the potential lower mutation rate. These findings provide insightful predictions about the potential future evolution of the enzyme, as well as plausible explanations for why these mutations have not yet appeared.

L162, positioned at the start of the Ω-loop and adjacent to the crucial catalytic residue E166, is assigned as the core residue for community 1 (Fig. 3A). While it remains conserved in SHV family, variants of L162 have been isolated in other class A β-lactamase and are known to expand the enzyme catalytic spectrum. Single amino acid substitution at L162 can intensify antibiotic resistance in BEL-1 (46), a class A ESBL clinical variant, exhibiting robust resistance to ticarcillin and ceftazidime (47). BEL-2 diverges from BEL-1 by single amino acid substitution (L162F) which alters the kinetic properties of the enzyme significantly and increases its affinity towards expanded-spectrum cephalosporins (48). The relationship between L162 and protein catalytic functions can be explained using DyNoPy model, as there are couplings with catalytic important residues M69, K73, E166, and K234. Moreover, the BEL case has confirmed that L162F mutation significantly destabilizes the overall protein structure, highlighting the crucial role of L162 in maintaining protein stability (46). DyNoPy accurately identifies the centrality of L162 by reporting its connections with 28 backbone residues, including nine hydrophobic node residues critical for protein stability. Among these, five hydrophobic residues are part of the α2 node: V75, L76, G78, V80, and L81, highlighting the contribution of L162 to the stability of the α2 helix (31).

Community 1, 4, and 5 of SHV-1 β-Lactamase.

All the residues are depicted as spheres on the protein structure. The core residue for each community is highlighted in red, while purple is used to emphasize the secondary core residue. Residues that interact with both cores are coloured in light yellow. Functional important residues are marked in cyan. Hydrophobic nodes are enclosed with cyan boxes. A. Community 1 of SHV-1, comprising 33 residues with L162 being the primary core residue. B. Community 4 of SHV-1, containing 12 residues and is centred by G156. C. Community 5 of SHV-1, embracing 48 residues and showing a strong correlation between V103 and S106.

Just like L162, N136 undergoes advantageous mutations in other class A β-lactamases while remains highly conserved within the SHV family. It is the core residue for community 7 (Fig. 4B). This residue forms a hydrogen bond with E166, stabilizing the Ω-loop (49). N170 acts as an intermediary between N136 and E166. N170, an essential catalytic residue located on the Ω-loop, contributes to priming the water molecule for the deacylation step with E166 (50) and is directly coupled with N136. Due to the essential contribution of N136 in facilitating E166 to maintain its proper orientation, it was previously thought to be intolerant to mutations as substitution of Asparagine to Alanine at this position would make the enzyme lose its function completely (51). However, N136D substitution has emerged as a new clinical variant very recently in PenL, a class A β-lactamase, by increasing its ability in hydrolysing ceftazidime (51), suggesting that this site has potential to mutate. This gain of function is mainly triggered by the increased flexibility of the Ω-loop (51). DyNoPy correctly detect a dynamical relationship between N136 and the Ω-loop (residues 164-179). Six residues present in the Ω-loop participate within this community, including R164 and D179. These two residues are critical as they are forming the ‘bottleneck’ of the Ω-loop which is essential for the correct position of E166 (52). D179 is also a critical mutation site for SHV-1. Single amino acid substitutions like D179A, D179N, and D179G are enough for the extended spectrum phenotype (42).

Community 6 and 7 of SHV-1 β-Lactamase.

All the residues are depicted as spheres on the protein structure. The core residue for each community is highlighted in red, while purple is used to emphasize the secondary core residue. Residues that interact with both cores are coloured in light yellow. Functional important residues are marked in cyan. A. Community 6 of SHV-1, comprising 30 residues with Y105 being the primary core residue. B. Community 7 of SHV-1, containing 34 residues and is centred by N136.

DyNoPy detects residue couplings essential for protein stability

DyNoPy identifies residue couplings critical for protein functional motions, particularly associated with protein stability. These residue pairs exhibit strong relationships as they are not only directly coupled with each other, but also forms various indirect couplings via other residues. As a result, both residues are considered as core residues inside these communities. It is expected that disruption of these couplings through mutation could compromise collective motions essential for enzyme activity.

As the secondary core residues in community 1 (Fig. 3A) F72 is showing a strong coupling with the primary core residue L162 and also forms nine indirect couplings with L162, including via the catalytic K234. This network of direct and indirect relationships reveals the importance of F72 and L162 coupling in maintaining protein functional motions. Interestingly, previous studies identified a small hydrophobic cavity formed by L162 and F72, together with L139, and L148, which is essential for the stability of the active site (46). Notably, DyNoPy successfully recovers the key residues of this local hydrophobic cavity (L162, F72, and L148).

The strong interplay between V103 and S106, which are both residues on the α3-α4 loop, is seen in community 5 (Fig. 3C). These residues not only interact with each other directly but are also indirectly coupled via 21 other residues. This community emphasizes the significance of hydrophobic nodes in SHV stability and dynamics. Within the analysed 48 residues, 27 are hydrophobic, out of which 15 residues act as nodes critical for enzyme stabilization. Hydrophobic nodes stabilize their own secondary structures and interconnect to stabilize the overall protein (53). V103 and S106 themselves are hydrophobic nodes, stabilizing α3 helix and α4 helix respectively, and are strongly coupled with each other. In CTX-M, another class A enzyme, N106S is a common substitution that results in improved thermodynamic stability and compensate for the loss in stability of the variants (54). Interestingly, this residue is already a Serine in SHV, but still implies its pivotal role in protein stability.

DyNoPy provides valid explanations for mutation sites

During the evolution of β-lactamases, single mutations on specific sites that are distant from the functional sites have been observed to significantly alter protein catalytic functions. Additionally, single mutations on some surface exposed residues can dramatically increase protein stability. Understanding how these distant mutations impact function and stability becomes a major challenge in understanding protein evolutionary pathways. Communities extracted by DyNoPy show these residues linked with functional important residues, providing a rational for these mutation sites with unknown functions.

Mutations of G156 are limited but they lead to ESBL phenotype in the SHV family (42). G156 is the central residue for community 4 (Fig. 3B), but it is distant from the active site, over 20Å away from the catalytic serine S70. Clinical variant SHV-27, has extended resistance ability towards cefotaxime, ceftazidime, and aztreonam (55). It differs from SHV-1 by single amino acid substitution G156D, suggesting that it has directly evolved from SHV-1 (55). Limited research has been done on position G156, and the understanding of how it affects the enzyme catalytic properties given that it is far away from the active site is still unclear. Based on our results, we suggest that this residue is essential for the overall protein function because of its 11 coevolved dynamic couplings with protein dynamics, including A146, another ESBL substitution site.

SHV-38, another ESBL that is capable of hydrolysing carbapenems, harbours a single A146V substitution compared to SHV-1 (56). Like G156, A146 is 15 Å away from S70 but shows an ability in altering protein catalytic function. The A146-G156 residue pair shows a strong coevolutionary signal and strong correlation with protein overall dynamics, implying that there may compensatory mutations at these sites with potential to emerge in the SHV family in the future. These two residues are not connected to any catalytic residues but their coupling to functional dynamics can offer plausible explanation to ESBL activity of these two mutations.

Unlike other substitution sites that are adjacent to the active site, R205 is situated more than 16 Å away from catalytic serine S70. Its side chain points outward from the protein, exposing to the solvent. The R205L substitution often co-occurs with other ESBL mutations and is thought to indirectly contribute to the ESBL phenotype by compensating for stability loss induced by other mutations (57). SHV-3 is an ESBL that exhibits significant resistance to cefotaxime and ceftriaxone (58). Two substitutions in this enzyme, R205L and G238S, extend its resistance profile (58). Thus, it is promising to see that DyNoPy detected these two mutation sites together within community 6 (Fig. 4A).

Y105 and R266 are the core residues for community 6. Y105 is situated on the α3-α4 loop positioned at the left side of the binding pocket. It is an important catalytic residue that recognizes and binds to the thiazolidine ring of penicillins or β-lactamase inhibitors (59). There is very limited information on the role of R266, except that it may stabilize the Ω-loop in the SHV family similar to the analogous T266 in TEM (60). G238 is coupled with an essential catalytic residue Y105, which further links with other catalytic functional residues: S70 and A237, and R266, a residue that known to stabilize the Ω-loop. This indicates that mutations on G238 would result in an alteration on protein catalytic function, as well as an increased flexibility of the protein, which strongly aligns with previous finding. Its linked mutation site R205 does not showing direct coupling with any catalytic residues. Instead, it is directly coupled with R266, which we mentioned as an Ω-loop stabilizer. Thus, it is not surprising that R205 substitution alone is never observed in nature (61), as it would not give significant evolutionary advantage to the protein.

Insights into unexplained functional sites of PDC-3

Unlike the extensively studied SHV-1, the functional roles of individual amino acids in PDC-3 remains largely unexplored. This gap in understanding serves as welcome challenge for interpreting the effects of mutations and the dynamic behaviour of PDC-3 from our results. Although several mutation hotspots, such as those on the Ω-loop (44), have been identified, very little is known about the specific contributions of individual amino acids on the functionality of PDC-3.

In PDC-3, mutations have primarily been reported in the Ω-loop. They enhance its flexibility to accommodate the bulky side chains of antibiotics, while deletions are more common in the R2-loop (39). DyNoPy detected five communities in total (Fig. S4B) with all the four predominant Ω-loop mutations appeared in these communities. Community 3, 4 and 5 are explained in the supporting information (Fig. S7). Furthermore, DyNoPy also detected several previously unexplored Ω-loop residues.

G214, a known mutation site in PDC-3, is the core residue in community 1. Another two essential mutation sites: E219 and Y221 also participate in this community, directly coupled with G214 (Fig. 5A). G214 also has direct couplings with four other Ω-loop residues: A195, A197, G212, and L216. Previous results have demonstrated that substitutions of Glycine to Alanine or Arginine at 214 significantly destabilizes the Ω-loop (32). The strong correlation between G214 and these Ω-loop residues emphasizes the significant contribution of G214 towards the stability of the Ω-loop, which corroborates with previous results (32). Moreover, substitutions such as G214A and G214R and mutations on E219 and Y221 do not affect R2 loop flexibility, resulting in the smaller active site volume among variants (32) because none of the residues from the R2 loop are detected in this community offering plausible explanation to previously unexplained phenomenon.

Community 1 and 2 of PDC-3 β-Lactamase.

All the residues are depicted as spheres on the protein structure. The core residue for each community is highlighted in red. Functional important residues are marked in cyan. A. Community 1 of PDC-3, comprising 36 residues with G214 being the primary core residue. B. Community 2 of PDC-3, containing 74 residues and is centred by G204.

G204 is the core residue of community 2, coupled with 73 other residues, most of which are distant from the catalytic site, suggesting plausible crucial role in overall protein stability like L162 in SHV-1 (Figure 5B). G204, a newly emerged mutation site in the PDC family (62), is located on the short β-sheet β5a within the Ω-loop, near the hinge region between β8 and β9 just above the active site. The only known variant of G204 is PDC-466, which was derived from PDC-462 (A89V, Q120K, V211A, N320S), with an addition of G204D (62). Coupling of G204 to several catalytically important residues, including K67, K315, and T316 can suggest that mutations at this site can negatively impact catalytic power. This offers a plausible explanation of seeing fewer variants at this site and mutations at this site could have impact on hydrolysing capabilities of PDC variants. This should be confirmed by further experimental studies of variants of G204. Unlike G214, E219 and Y221 mutations which do not influence the dynamics of the R2 loop, substitutions on V211, a member of Ω-loop, has impact on dynamics of R2 loop because of its indirect couplings, through G204 to R2-loop residues (32). Two less critical substitution sites, H188 and V329, were also observed in community 2.

Conclusions

DyNoPy combines information on residue-residue coevolution and protein dynamics captured from MD ensembles to detect conserved dynamic couplings. These couplings are modelled as a graph. The network analysis facilitates the extraction of epistatic subnetworks and the assignment of roles to residues based on their importance in the graph model. The choice of a relevant descriptor of protein dynamics has an impact on the ability to detect couplings that are involved in functional dynamics. Here we demonstrated how the choice of relevant global and local descriptors returns a higher number of effective couplings (greater than 0), and in turn leads to interpretable graph models and communities. In other systems, when multiple descriptors can be used to quantify functional conformational change, it is expected that they will differently modulate the effect of coevolution coupling, which will be reflected in a different structure of the associated graph models. This suggests the use of DyNoPy to generate comparative models in proteins with multiple functions associated to distinct dynamical changes.

Mutations of L162 and N136 have not yet emerged in SHV-1, but they are detected by DyNoPy as core residues for communities. These residues are strongly coupled with other functional important residues, which play critical roles in protein stability and catalytic activity. The identification of these couplings shows high consistency with previous studies and highlights the importance of L162 and N136 in SHV-1 functional dynamics. Given their central role in these communities, mutations in L162 and N136 can significantly alter protein function, suggesting their potential for future evolutionary changes. However, their strong relationships with these critical functional residues also suggest that mutation at these sites would need to be balanced to maintain protein function, providing an explanation for why such mutations have not yet emerged in SHV-1 (63). The ability of DyNoPy in detecting functionally important mutation sites was demonstrated via well-characterized mutation sites including R205 and G238 from SHV-1. Moreover, DyNoPy shows predictive ability on less-studied mutation sites such as G156 and A146, by detecting critical residue couplings that coevolved with functional motions.

Based on the knowledge we have gained from analysis of SHV-1 functional protein dynamics we suggest that in PDC-3, mutations at G204 because of its significant conserved dynamic couplings can lead to new ESBL/IRBL clinical variants. We suggest that DyNoPy can be used as a predictive tool to identify potential functional residues within this enzyme and guide future mutagenesis studies.

In summary, by integrating hidden evolutionary information with direct dynamic interactions, DyNoPy provides a powerful framework for identifying and analysing functional sites in proteins. The tool not only identifies key residues involved in local and global interactions, but also improves our ability to predict silent residues with previously unknown roles for future experimental testing. Our application of DyNoPy to broad-spectrum β-lactamases ESBLs and IRBLs demonstrates its potential to address key medical challenges such as antibiotic resistance by providing valid predictions on protein evolution.

Methodology

DyNoPy generates a graph representation of the protein structure that captures the couplings between amino acid residues contributing to the functional dynamics of the protein. Residues are represented as graph nodes, and conserved dynamic couplings are recorded as edges. Edge weights quantify the strength of these couplings. The model is built on two assumptions: residue pairs should have i) coevolved and their ii) time-dependent interactions correlate with a functional conformational change.

Therefore, edge weights (Jij) for residue i and j are calculated as: where γij is the scaled coevolution score and ρij is the degree of correlation with the selected functional conformational change. α and β are weights assigned to γij and ρij that have a sum of one. The relative weight of the scaled coevolution score (α) is set to 0.5 in this study. When either of the assumptions listed above is not met, Jij is set to zero.

Scaled coevolution scores

The occurrence of residue-residue coevolution can be estimated and quantified using probabilistic models of correlated mutations from deep multiple sequence alignments (MSA). DyNoPy supports generation of the MSA using the HH-Suite package (64) and calculation of scaled coevolution score (γij) using CCMpred (65). First a pairwise residue coevolution matrix (C) is calculated, then these raw scores (cij) are divided by the matrix mean (Equation 2). All scores (sij) smaller than 1 are set to zero, and the remaining values are normalised by the maximum value (Equation 3):

Correlation with functional motions

The contribution of a residue pair to a selected functional motion is estimated by how much the change in interaction energy between the two residues over time is correlated with a collective variable (CV) describing the functional motion: where εij(t) is the pairwise non-bonded interaction energy (see details in Supporting Information) and d(t) is the time-dependent value of the CV. Examples of CV and a discussion on the choice of the most relevant CV is presented in the results section. Correlation values smaller than 0.5 are set to 0.

Graph representation and analysis of conserved dynamic couplings

All pairwise conserved dynamic couplings (Equation 1) are collected into a square matrix J. A graph is built from J, using python-igraph v0.11 library (66). Nodes represent residues, and edges are drawn between nodes with positive Jij. Edge weights are set to Jij. The relative importance of the residues in this model of protein dynamics is calculated as eigenvector centrality of the nodes (67). Residues involved in extensive correlated dynamics with other highly connected residues have higher eigenvector centrality (EVC) scores. Groups of residues contributing to important collective motions are detected by community analysis of the graph structure. The Girvan-Newman algorithm is used to extract the community structure (68).

Adaptive Sampling Molecular Dynamics Simulations

MD simulation data was sourced from our previous studies (31, 32). To summarise, SHV-1 structural coordinates (PDB ID: 3N4I) were obtained from the Protein Data Bank and modified to the wild type by introducing the E104D mutation. Similarly, the PDC-3 structure was derived from PDC-1 (PDB ID: 4HEF) by a T105A substitution. Both enzymes were protonated at pH 7.0 using PropKa from the PlayMolecule platform (69). One disulfide bond between C77 and C123 was specified in SHV-1. Both structures were solvated with TIP3P water molecules in a periodic box with a box size of 10 Å. Ions were added to neutralize the overall charge of each system at 150mM KCl. Amber force field ff14SB was used for all MD simulations (70). After an initial minimisation of 1000 steps, both the enzymes were equilibrated for 5 ns in the NPT ensemble at 1 atmospheric pressure using the Berendsen barostat (71). The initial velocities for each simulation were sampled from the Boltzmann distribution at 300 K. Multiple Markov State Model (MSM)-based adaptively sampled simulations were performed for both proteins based on the ACEMD engine (72, 73). A canonical (NVT) ensemble with a Langevin thermostat (74) (damping coefficient of 0.1 ps−1) and a hydrogen mass repartitioning scheme were employed to achieve time steps of 4 fs. For SHV-1, each trajectory spanned 60 ns with a time step of 0.1 ns, with a total of 593 trajectories. In the case of PDC-3, 100 trajectories were collected, each containing 3000 frames, lasting 300 ns. To manage the extensive datasets efficiently, trajectories were strategically stridden to ensure that a minimum of 30,000 frames were preserved for each system. The resulting trajectories are summarized in Supporting Table S4.

Calculation and Selection of Collective Variables

Any vector that yields time-dependent information can be used as a collected variable (CV) to describe protein functional motions. The usefulness of DyNoPy is dependent on the choice of the CVs. To guide the selection of CVs, we selected 12 distinct features: radius of gyration (Rg), the first principal component (PC1), partial PC1 (PC1_partial), the first time-lagged independent component (TC1), partial TC1 (TC1_partial), global root mean square deviation (gRMSD), partial RMSD (pRMSD), dynamical RMSD (dRMSD), global solvent accessible surface area (gSASA), partial SASA (pSASA), active site pocket volume, and the number of hydrogen bonds (hbond). A description of the CVs, including the calculation methods and the residues used to calculate the partial variables, is detailed in the supporting information. CVs were subsequently used as input features for DyNoPy. The selection criteria for CVs were based on the number of residue pairs each CV could detect.

Data Availability

All files required to run the simulations (topology, coordinates, input), processed trajectories (xtc), corresponding coordinates (pdb), can be downloaded from the DOI https://doi.org/10.57760/sciencedb.15876 (PDC-3) and 10.5281/zenodo.13693144 (SHV-1). DyNoPy is available at https://github.com/alepandini/DyNoPy.

eLife Assessment

This paper reports the analysis of coevolutionary patterns and dynamical information for identifying functionally relevant sites. These findings are considered important due to the broad utility of the unified framework and network analysis capable of revealing communities of key residues that go beyond the residue-pair concept. The data is solid, and the results are clearly presented.

Reviewer #1 (Public review):

Summary:

As reported above, this paper by Xu et al reports on a new method to combine the analysis of coevolutionary patterns with dynamic profiles to identify functionally important residues and reveal correlations between binding sites.

Strengths:

In general, coevolutionary analysis and MD analysis are carried out separately and while there have been attempts to compare the information provided by the two, no unified framework exists. Here, the authors convincingly demonstrate that integrating signals from Dynamics and coevolution gives information that substantially overcomes the one provided by either method in isolation. While other methods are useful, they do not capture how dynamics is fundamental to define function and thus sculpts coevolution, via the 3D structure of the protein. At the same time, the authors demonstrate how coevolution in turn also influences internal dynamics. The Networks they rebuild unveil information at an even higher level: the model starts pairwise but through network representation the authors arrive to community analysis, reporting on interaction patterns that are larger than simple couples.

Weaknesses:

The authors should

-Make an effort in suggesting/commenting the limits of applicability of their method;

-Expand discussion on how DyNoPy compares to other methods;

-Dynamic is not essential in all systems (structural proteins): The authors may want to comment on possible strategies they would use for other systems where their framework may not be suitable/applicable.

Reviewer #2 (Public review):

Summary:

Authors introduced a computational framework, DyNoPy, that integrates residue coevolution analysis with molecular dynamics (MD) simulations to identify functionally important residues in proteins. DyNoPy identifies key residues and residue-residue coupling to generate an interaction graph and attempts to validate using two clinically relevant β-lactamases (SHV-1 and PDC-3).

Strengths:

DyNoPy could not only show clinically relevance of mutations but also predict new potential evolutionary mutations. Authors have provided biologically relevant insights into protein dynamics which can have potential applications in drug discovery and understanding molecular evolution.

Weaknesses:

Although DyNoPy could show the relevance of key residues in active and non-active site residues, no experiments have been performed to validate their predictions. In addition, they should compare their method with conventional techniques and show how their method could be different.

An explanation of "communities" divided in the work and how these communities are relevant to the article should be provided. In addition, choice of collective variables and their relevance in residue coupling movement is also not very well explained. Dynamics cross correlation map can also be a good method for understanding the residue movements and can explain the residue-residue coupling, it is not explained how DyNoPy is different from the conventional methods or can perform better.

In the sentence "DyNoPy identified eight significant communities of strongly coupled residues within SHV-1 (Supporting Fig. S4A)" I could not find a clear description of eight significant communities.

Again the description of communities is not clear to me in the following sentence "Detailed description of the other three communities is provided in the supporting information (Fig. S6)."

In the sentence "N170 acts as an intermediary between N136 and E166". Kindly cite the reference figure to show N179 as intermediate residue.

Please be careful with the numbers. In the sentence "These residues not only interact with each other directly but are also indirectly coupled via 21 other residues." I could count 22 other residues and not 21.

In the sentence "Unlike other substitution sites that are adjacent to the active site, R205 is situated more than 16 Å away from catalytic serine S70". Please add this label somewhere in the figure.

Please cite a reference in the sentence "This indicates that mutations on G238 would result in an alteration on protein catalytic function, as well as an increased flexibility of the protein, which strongly aligns with previous finding."

Reviewer #3 (Public review):

Summary:

In this paper, Xu, Dantu and coworkers report a protocol for analyzing coevolutionary and dynamical information to identify a subset of communities that capture functionally relevant sites in beta-lactamases.

Strengths:

The combination of coevolutionary information and metrics from MD simulations is interesting for capturing functionally relevant sites, which can have implications in the fields of drug discovery but also in protein design.

Weaknesses:

The combination of coevolutionary information and metrics from MD simulations is not new as other protocols have been proposed along the years (the current version of the paper neglects some of them, see below), and there are a few parameters of the protocol that, in my opinion, should be better analyzed and discussed.

(1) As mentioned, the introduction of the paper lacks some important publications in the field of using graph theory to represent important interaction networks extracted from MD simulations (DOI: 10.1002/pro.4911), and also combining MD data with MSA to identify functionally relevant sites for enzyme design (doi: 10.1021/acscatal.4c04587, 10.1093/protein/gzae005).

(2) The matrix used to apply graph theory (J_ij) is built from summing the scaled coevolution and degree of correlation values. The alpha and beta weights are defined, and the authors mention that alpha is set to 0.5, thus beta as well to fulfil with the alpha + beta = 1. Why a value of 0.5 has been selected? How this affects the overall results and conclusions extracted? The finding that many catalytically relevant residues are identified in the communities is not surprising given that such sites usually present a high conservation score.

(3) Another important point that needs further explanation is the selection of the relevant descriptor of protein dynamics. In this study two different strategies have been used (one more global the other more local), but more details should be provided regarding their choice. What is the best strategy according to the authors? Why not using the same strategy for both related systems? The obtained results using one methodology or the other will have a large impact on the dynamical score. Another related point is: what is the impact of the MD simulation length, how the MSA is generated and number of sequences used for MSA construction?

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

As reported above, this paper by Xu et al reports on a new method to combine the analysis of coevolutionary patterns with dynamic profiles to identify functionally important residues and reveal correlations between binding sites.

Strengths:

In general, coevolutionary analysis and MD analysis are carried out separately and while there have been attempts to compare the information provided by the two, no unified framework exists. Here, the authors convincingly demonstrate that integrating signals from Dynamics and coevolution gives information that substantially overcomes the one provided by either method in isolation. While other methods are useful, they do not capture how dynamics is fundamental to define function and thus sculpts coevolution, via the 3D structure of the protein. At the same time, the authors demonstrate how coevolution in turn also influences internal dynamics. The Networks they rebuild unveil information at an even higher level: the model starts pairwise but through network representation the authors arrive to community analysis, reporting on interaction patterns that are larger than simple couples.

Weaknesses:

The authors should

- Make an effort in suggesting/commenting the limits of applicability of their method;

We have added a sentence on Page 17, line 15 that describes the limitation of our method.

- Expand discussion on how DyNoPy compares to other methods;

A paragraph has been added to explain the comparison with other models (Page 3, line 18)

- Dynamic is not essential in all systems (structural proteins): The authors may want to comment on possible strategies they would use for other systems where their framework may not be suitable/applicable.

We agree with the reviewer that dynamics is not essential in all systems. In systems where there is limited role of dynamics in the function, the analysis done with DyNoPy is equivalent to conventional coevolution analysis, which can be consider one limitation of our method. Conversely, for dynamic proteins, combining functional dynamics descriptors with coevolution analysis using DyNoPy, helps in denoising information by deconvolution of communities. We have included this in the manuscript to highlight the suitability/applicability of the method.

Further, we have added a paragraph in the Introduction and conclusions highlighting the main difference between DyNoPy and existing computational tools like DCCM, KIN, and SPM and for your convenience it is provided below:

“Functional sites are often regulated by both, local and global interactions. Changes in these interactions are instrumental for functional events like substrate binding, catalysis, and conformational changes (18). The development of physical models of protein dynamics and the increase in available computational power has stimulated the adoption of computational techniques (19, 20) to investigate the conformational dynamics of proteins, an essential component of the many biological functions (21, 22). Different models have been proposed to describe the interactions between residues during simulations and network models have been particularly popular, including methods on single structures and MD simulations data built by analysing the response to external forces on residue networks (23), by estimating the prevalence of non-covalent energy interaction networks in homologous proteins (24), or by analysing linear or non-linear correlation in atomic fluctuations (25, 26). These techniques have demonstrated their usefulness in extracting allosteric networks from structural data with applications in enzyme design (26).”

Reviewer #2 (Public review):

Summary:

Authors introduced a computational framework, DyNoPy, that integrates residue coevolution analysis with molecular dynamics (MD) simulations to identify functionally important residues in proteins. DyNoPy identifies key residues and residue-residue coupling to generate an interaction graph and attempts to validate using two clinically relevant β-lactamases (SHV-1 and PDC-3).

Strengths:

DyNoPy could not only show clinically relevance of mutations but also predict new potential evolutionary mutations. Authors have provided biologically relevant insights into protein dynamics which can have potential applications in drug discovery and understanding molecular evolution.

Weaknesses:

Although DyNoPy could show the relevance of key residues in active and non-active site residues, no experiments have been performed to validate their predictions.

We thank the reviewer for highlighting this point. We acknowledge that direct experimental validation of our predictions for DyNoPy has not yet been performed. However, we have provided explanations and evidence from experiments conducted on closely related homologs to support the relevance of key residues. These homologs share significant structural and functional similarity, which strengthens the reliability of our predictions.

In addition, they should compare their method with conventional techniques and show how their method could be different.

We thank all the reviewers for highlighting this oversight on our behalf. In Introduction and conclusion, we have added the following paragraphs:

“Functional sites are often regulated by both, local and global interactions. Changes in these interactions are instrumental for functional events like substrate binding, catalysis, and conformational changes (18). The development of physical models of protein dynamics and the increase in available computational power has stimulated the adoption of computational techniques (19, 20) to investigate the conformational dynamics of proteins, an essential component of the many biological functions (21, 22). Different models have been proposed to describe the interactions between residues during simulations and network models have been particularly popular, including methods on single structures and MD simulations data built by analysing the response to external forces on residue networks (23), by estimating the prevalence of non-covalent energy interaction networks in homologous proteins (24), or by analysing linear or non-linear correlation in atomic fluctuations (25, 26). These techniques have demonstrated their usefulness in extracting allosteric networks from structural data with applications in enzyme design (26). ”

An explanation of "communities" divided in the work and how these communities are relevant to the article should be provided. In addition, choice of collective variables and their relevance in residue coupling movement is also not very well explained. Dynamics cross correlation map can also be a good method for understanding the residue movements and can explain the residue-residue coupling, it is not explained how DyNoPy is different from the conventional methods or can perform better.

The following sentences have been included in the manuscript to address the questions raised by the reviewer:

On Community Definition and relevance

DyNoPy identified coevolving residue pairs (scaled coevolution score >1) with interactions strongly correlated with protein functional motions (i.e., J values larger than zero). Applying network analysis on the combined dynamics-coevolution matrix helps us extracting higher-order interactions beyond pairwise coupling and detecting critical residues, which show multiple interactions with each other. Moreover, indirect long-range relationships, which would be hard to identify from numerical data, could be detected through community clustering. Community-based analysis offers a more comprehensive understanding of residue relationships and enables the visualization of residue couplings on the protein structure.

On Choice of collective variables:

DyNoPy works on the assumption that time-dependent interactions between critical residues, either having significant structural change or not will correlate with functional conformational motions. Since MD simulation data is high-dimensional, a time-dependent dynamic descriptor is required to extract the most relevant information for the process under study. A good collective variable (CV) should appropriately describe protein functional motions. Thus, a CV that detects the highest number of residue couplings is expected to be the most suitable descriptor (Mentioned in Page 22 Line 14). In our study, we tested 12 CVs, either focusing on the entire protein or on selected regions. And the best performed CV (the one identified the most residue couplings) was selected for further analysis. In practical applications, users can decide whether to focus on the most relevant global or local dynamics descriptor depending on the dynamics of their specific system.

We have added a paragraph in the Introduction differentiating DyNoPy with other methods including DCCM. DCCM differs from DyNoPy in two aspects 1) it does not account for inter-residue coevolution 2) the correlation matrix captures correlations of atomic/residue movements associated with the whole intrinsic dynamics of the system, without filtering for the contributions to the important motions involved in the biological function. Additionally, any residue pair contributing to functional motion without itself undergoing any structural change will not be visible in this approach.

In the sentence "DyNoPy identified eight significant communities of strongly coupled residues within SHV-1 (Supporting Fig. S4A)" I could not find a clear description of eight significant communities.

The following sentences have been included in the results, methods and figure legends that define ‘significant community’:

‘DyNoPy identified eight meaningful communities, each consisting of at least three strongly coupled residues within SHV-1 (Supplementary Fig. S4A). All crucial catalytic residues and critical substitution sites previously mentioned participating in one of these communities with the exceptions of R43, R202, and S130.’ (Page 8 Line 28)

‘A meaningful community should contain at least three residues.’ (Page 21 Line 2)

‘A reasonable residue community should contain at least three residues.’ (SI Page 11)

Again the description of communities is not clear to me in the following sentence "Detailed description of the other three communities is provided in the supporting information (Fig. S6)."

This following sentence has been rewritten.

‘Detailed description of communities with secondary importance for protein function (community 3, 8, and 9) is provided in the supplementary information (Supplementary Fig. S6).’ (Page 9, line 8)

In the sentence "N170 acts as an intermediary between N136 and E166". Kindly cite the reference figure to show N179 as intermediate residue.

This sentence has been rewritten to avoid any confusion.

‘Although DyNoPy did not detect this direct interaction between N136 and E166, the established relationship between N136 and N170 highlights the role of N136 in influencing E166.’ (Page 10 Line 8)

Please be careful with the numbers. In the sentence "These residues not only interact with each other directly but are also indirectly coupled via 21 other residues." I could count 22 other residues and not 21.

We thank the reviewer for spotting this error. This has now been corrected. All the communities are counted again.

‘These residues not only interact with each other directly but are also indirectly coupled via 22 other residues.’ (Page 12 Line 14)

In the sentence "Unlike other substitution sites that are adjacent to the active site, R205 is situated more than 16 Å away from catalytic serine S70". Please add this label somewhere in the figure.

The figure legends have been updated to include this. Distances have been added to community 4 Fig. 3 and community 6 Fig. 4. Residue index in the legend of Fig.3 has been included as subscript. Distance in the main text has been changed to be more accurate.

‘G156 and A146 are two functional important residues distant from the active site. G156 is 21.3Å away from the catalytic S70. A146 is 16.8Å away from S70.’ (Page 12 Line 2)

‘R205 is a functional important residue that is 20.6Å away from the active site S70.’ (Page 13 Line 10)

Please cite a reference in the sentence "This indicates that mutations on G238 would result in an alteration on protein catalytic function, as well as an increased flexibility of the protein, which strongly aligns with previous finding."

The citation has been added

‘This indicates that mutations on G238 would result in an alteration on protein catalytic function, as well as an increased flexibility of the protein, which strongly aligns with previous finding (62).’ (Page 15 Line 2)

Reviewer #3 (Public review):

Summary:

In this paper, Xu, Dantu and coworkers report a protocol for analyzing coevolutionary and dynamical information to identify a subset of communities that capture functionally relevant sites in beta-lactamases.

Strengths:

The combination of coevolutionary information and metrics from MD simulations is interesting for capturing functionally relevant sites, which can have implications in the fields of drug discovery but also in protein design.

Weaknesses:

The combination of coevolutionary information and metrics from MD simulations is not new as other protocols have been proposed along the years (the current version of the paper neglects some of them, see below), and there are a few parameters of the protocol that, in my opinion, should be better analyzed and discussed.

(1) As mentioned, the introduction of the paper lacks some important publications in the field of using graph theory to represent important interaction networks extracted from MD simulations (DOI: 10.1002/pro.4911), and also combining MD data with MSA to identify functionally relevant sites for enzyme design (doi: 10.1021/acscatal.4c04587, 10.1093/protein/gzae005).

We are very grateful for pointing us to these references. We have added a paragraph in the Introduction mentioning these and other computational tools similar to DyNoPy. Further, in conclusion we have highlighted the differences between DyNoPy and existing tools.

(2) The matrix used to apply graph theory (J_ij) is built from summing the scaled coevolution and degree of correlation values. The alpha and beta weights are defined, and the authors mention that alpha is set to 0.5, thus beta as well to fulfil with the alpha + beta = 1. Why a value of 0.5 has been selected? How this affects the overall results and conclusions extracted? The finding that many catalytically relevant residues are identified in the communities is not surprising given that such sites usually present a high conservation score.

This is an excellent question. Our present formulation allows the user to easily assess the influence of coevolution and dynamic couplings on the output. Setting alpha to 0.5, weights both evolutionary and dynamics information equally and has shown promising results in SHV-1 and PDC-3. As it has been presented in the manuscript, setting alpha to 1, i.e., purely utilising coevolution data does not let us identify critical residues effectively as all residues are included in the set (Supplementary Fig. S4 and S5). In future work, we would like to investigate the effect of scanning alpha from 0 to 1 on the final residue list, possibly on a larger set of proteins and protein families.

We would also like to point out that some of the residue pairs with coevolution scores in the top 1% have J-scores set to 0, as they lacked significant coupling to the functional dynamics.

(3) Another important point that needs further explanation is the selection of the relevant descriptor of protein dynamics. In this study two different strategies have been used (one more global the other more local), but more details should be provided regarding their choice. What is the best strategy according to the authors? Why not using the same strategy for both related systems? The obtained results using one methodology or the other will have a large impact on the dynamical score. Another related point is: what is the impact of the MD simulation length, how the MSA is generated and number of sequences used for MSA construction?

As in the case of many complex proteins, the flow of information occurs in β-lactamases via structural interactions (https://doi.org/10.7554/eLife.66567). These interactions occur both on a local level, as in the case of binding site residues or residues immediately surrounding the binding site; however, there are interactions far away (>20Å) from the binding site that have the ability to alter function. We have obtained this information from extensive surveys of clinical isolates and experimental data. To account for such interactions, a more global approach has to be taken. To answer the reviewer’s question: each system is unique and there is no one-fixed strategy. In short, the method used should be able to denoise information and the user is advised to fine-tune their findings by corroborating with experimental and clinical information.

The length of MD simulations is also system specific. Some systems effectively sample the functional dynamics within a shorter simulation time, while others take a long timescale MD simulation to converge. The results won’t change as long as the simulation has effectively sampled the functional dynamics associated with biological function.

The MSA is generated by the HH-Suite package as mentioned on Page 19 Line 19. More specifically, the MSA is constructed based on the UniRef30 database, where sequences are clustered, and each cluster contains sequences with at least 30% sequence identity. This provides a non-redundant set of protein sequences. Our package allows the automatic generation of MSAs from the database. For SHV-1, the alignment contains 18,175 protein sequences and for PDC-3, the alignment consists of 27,892 protein sequences. Full details of this protocol are published in Bibik et al. (https://doi.org/10.1093/bioinformatics/btae166). We have revised the methods section to include these details.

Other Minor Alterations

‘Fig. S1 and S2’ has been changed to ‘Supplementary Fig. S1 and S2’ for consistency (Page 6 Line 12)

(1) ‘Figure 5B’ has been changed to ‘Fig. 5B’ for consistency (Page 16 Line 11)

(2) All the ‘Figure’ has been changed to ‘Fig.’ in the SI for consistency

(3) Just as the suggestion, an alteration has been made on the Step 1 of Fig.1.

Version 2

Abstract

Abstract

The relationship between protein dynamics and function is essential for understanding biological processes and developing effective therapeutics. Functional sites within proteins are critical for activities such as substrate binding, catalysis, and structural changes. Existing computational methods for the predictions of functional residues are trained on sequence, structural and experimental data, but they do not explicitly model the influence of evolution on protein dynamics. This overlooked contribution is essential as it is known that evolution can fine tune protein dynamics through compensatory mutations, either to improve the proteins’ performance or diversify its function while maintaining the same structural scaffold. To model this critical contribution, we introduce DyNoPy, a computational method that combines residue coevolution analysis with molecular dynamics (MD) simulations, revealing hidden correlations between functional sites. DyNoPy constructs a graph model of residue-residue interactions, identifies communities of key residue groups and annotates critical sites based on their roles. By leveraging the concept of coevolved dynamical couplings—residue pairs with critical dynamical interactions that have been preserved during evolution—DyNoPy offers a powerful method for predicting and analysing protein evolution and dynamics. We demonstrate the effectiveness of DyNoPy on SHV-1 and PDC-3, chromosomally encoded β-lactamases linked to antibiotic resistance, highlighting its potential to inform drug design and address pressing healthcare challenges.

Introduction

Quantifying the contribution of individual residues or residue groups to protein function is important to estimate the pathogenic effect of mutations (1). Identifying the functional roles of individual residues has primarily been done through mutagenesis experiments (2). Bioinformatics methods have complemented these approaches through analysis of multiple sequence alignments (MSA) of homologous proteins and structural data (3–8). Among these methods, computational techniques that can decode inter-residue evolutionary relationships from MSAs have paved the way for machine learning (ML) based strategies that can predict protein structure (9–12), stability (13), and function (7) and extend the scope of computational protein design (14–16). A most recent approach has combined experimental data from three proteins, NUDT15, PTEN and CYP2C9, on stability and function with sequence and structural features to train a ML model to predict functional sites (17).

Functional sites are often regulated by both, local and global interactions. Changes in these interactions are instrumental for functional events like substrate binding, catalysis, and conformational changes (18). The development of physical models of protein dynamics and the increase in available computational power has stimulated the adoption of computational techniques (19, 20) to investigate the conformational dynamics of proteins, an essential component of the many biological functions (21, 22). Different models have been proposed to describe the interactions between residues during simulations and network models have been particularly popular, including methods on single structures and MD simulations data built by analysing the response to external forces on residue networks (23), by estimating the prevalence of non-covalent energy interaction networks in homologous proteins (24), or by analysing linear or non-linear correlation in atomic fluctuations (25, 26). These techniques have demonstrated their usefulness in extracting allosteric networks from structural data with applications in enzyme design (26).

However, none of these techniques incorporate information on residue evolution into the computational approach, While it has been established that evolution through compensatory mutations in dynamic regions, like hinges and loops, can fine tune protein structural dynamics and introduce promiscuity, thereby diversifying biological function. Assuming that protein functional dynamics is conserved during evolution, significant information on dynamic regions and substrate recognition sites should be recoverable using inter residue coevolution scores extracted from MSAs (27, 28). Coevolution analysis and Molecular Dynamics (MD) simulations have independently (29) and synergistically been combined in the past to identify important residues for function (30–34). Yet a method that combines hidden information on dynamics from evolution with direct information on local and global dynamics from conformational ensembles from MD is not yet available.

Here, we present DyNoPy, a computational method that can extract hidden information on functional sites from the combination of pairwise residue coevolution data and powerful descriptors of dynamics extracted from the analysis of MD ensembles. The method can detect coevolved dynamic couplings, i.e. residue pairs with critical dynamical interactions that have been preserved during evolution. These pairs are extracted from a graph model of residue-residue interactions. Communities of important residue groups are detected, and critical sites are identified by their eigenvector centrality in the graph (Fig. 1). We demonstrate the power of this approach on SHV-1 and PDC-3 β-lactamases of major clinical importance (35, 36). DyNoPy successfully detects residue couplings that align with previous studies, guide in the explanations of mutation sites with previously unexplained mechanisms and provide predictions on plausible important sites for the emergence of clinically relevant variants.

Overview of the DyNoPy workflow

Results and Discussion

β-lactamases are a group of enzymes capable of hydrolysing β-lactams, conferring resistance to β-lactam antibiotics (37). These enzymes are evolving rapidly, as single amino acid substitutions are sufficient to drive their evolution and increase their catalytic spectrum and inhibitor resistance profile (38). The widespread dissemination of β-lactamases across different bacterial species and their extensive emergence highlight their global impact on antibiotic resistance (39). The rapid evolution of β-lactamases and their clinical significance (38) makes them an ideal target for evaluating the robustness of DyNoPy.

In this study, we applied DyNoPy to two model enzymes from different β-lactamase families: class A β-lactamase SHV-1 (a chromosomally encoded enzyme in Klebsiella pneumoniae) and class C β-lactamase PDC-3 (a chromosomally encoded enzyme in Pseudomonas aeruginosa) (35, 36) (Supplementary Fig. S1 and S2). Both class A and class C β-lactamases comprise an α/β domain and an α helical domain, with the active site situated in between (40, 41). Moreover, both enzymes target the carbonyl carbon of the β-lactams using a highly conserved serine residue (42, 43). Despite these similarities, the structures of class A and class C β-lactamases are remarkably different (Fig. 2).

Structural Comparison of SHV-1 (PDB ID: 3N4I) and PDC-3 (PDB ID: 4HEF) β-Lactamases. Catalytic serine S70 (SHV-1) and S64 (PDC3) are highlighted using stick representation. Important loops surrounding the active site are highlighted in red. In SHV-1, highlighted loops are the α3-α4 loop (residues 101-111), the Ω-loop (residues 164-179), and the hinge region (residues 213-218). In PDC-3, highlighted loops are the Ω-loop (residues 183-226) and the R2-loop (residues 280-310).

SHV-1 is a very well characterised enzyme with wealth of information on mutations and their corresponding effects on protein function. In contrast, the information available on PDC-3 remains limited. Detailed structural information on these enzymes can be found in the supplementary materials. Essential catalytic residues in SHV-1 are: S70, K73, S130, E166, N170, K234, G236, and A237 (44) and conserved catalytic residues in PDC-3 include S64, K67, Y150, N152, K315, T316, and G317. Highly conserved stretches of 3-9 hydrophobic residues, annotated as hydrophobic nodes, exists in class A β-lactamases and have been proven to be essential for protein stability (45). Residues defined as belonging to hydrophobic nodes within SHV-1 are listed in Supplementary Table S1.

In SHV-1, the predominant extended spectrum β-lactamase (ESBL) substitutions occur at L35, G238, and E240, while R43, E64, D104, A146, G156, D179, R202, and R205 appear in ESBLs with lower frequency (46). Mutations at M69, S130, A187, T235, and R244 are known to induce inhibitor resistance in the enzyme (47). In PDC-3, substitutions primarily occur on the Ω-loop, enhancing its flexibility to accommodate the bulky side chains of antibiotics, while deletions are more common in the R2-loop (43). The predominant Ω-loop mutations isolated from clinics are found at positions V211, G214, E219, and Y221 (48).

Emergence of highly conserved dynamic couplings

DyNoPy builds a pairwise model of conserved dynamic couplings detected by combining coevolution scores and information on functional motions into a score Jij (see Methods and Fig. 1). To this end a dynamics descriptor should be selected. When the descriptor is associated with functional conformational changes, it is expected that functionally relevant couplings will report higher scores. Dynamics descriptors can be selected from commonly used geometrical collective variables (CVs) for the analysis of MD trajectories (see Methods). As expected, the average J matrix score varies across the different CVs, with some of them showing no signal of dynamic coupling (Supplementary Fig. S4C).

SHV-1 and PDC-3 exhibit distinct dynamics, requiring a different choice of the CV that best captures the functional dynamics. For SHV-1, the global first principal component (PC1) proved to be the most effective feature, identifying 571 residue pairs with a Jij value greater than 0. Conversely, PDC-3 requires selection of more localized features that can extract the Ω-loop dynamics from the overall protein motion. Among the dynamic descriptors, the partial first time-lagged component (TC1_partial) performed best for PDC-3, detecting 216 residue pairs with a Jij value greater than 0. Consequently, PC1 and TC1_partial were selected to build the J matrix for SHV-1 and PDC-3, respectively. The performance of all 12 CVs for each protein was assessed and listed in the Supplementary Table S2.

The importance of dynamical information is evident when coevolution couplings (γij) and conserved dynamic couplings (Jij) are compared: the number of non-zero couplings decrease from 40% to <2% of total residue pairs in the protein (Supplementary Fig. S4D) when information from the dynamics descriptor is added. Thus, the inclusion of protein dynamics in coevolution studies acts as an effective filter that rules out residue pairs that do not have significant correlations with functional motions. Moreover, when relying only on γij, all the residues in SHV-1 and PDC-3 are included within four identified communities (Supplementary Table S3), suggesting that coevolution scores (γij) alone do not effectively discriminate residues relevant for protein functions. Furthermore, it would be hard to distinguish critical core residues for each community using only γij, as the eigenvector centrality (EVC) values for the residues do not show remarkable differences (Supplementary Fig. S5A and S5B). This means that detailed dynamic investigation of the top residues is needed to determine which pairs should be picked up and further analysed. On the other hand, it is much easier to identify essential residues based on J scores calculated, as clear outliers with significantly higher EVC values could be seen for almost all communities (Supplementary Fig. S5C and S5D) (29, 49). In conclusion, the lack of specificity in the statistically based coevolution analysis supports the choice of incorporating a score for the correlation between residue interactions and dynamic behaviours that enables deconvolution of community information.

DyNoPy reveals critical residues and predicts evolutionary pathways in SHV-1

DyNoPy identified eight meaningful communities, each consisting of at least three strongly coupled residues within SHV-1 (Supplementary Fig. S4A). All crucial catalytic residues and critical substitution sites previously mentioned participating in one of these communities with the exceptions of R43, R202, and S130. Residues previously known to have critical role in function or conferring ESBLs/IRBLs phenotype are either directly coupled to protein dynamics or act as a central hub. The hubs interact with residues with either a role in catalysis or structural stability through their membership of hydrophobic nodes (35). Furthermore, DyNoPy identified key positions (L162 and N136) within some communities that are known to undergo substitutions, conferring an ESBL phenotype in other class A β-lactamases. These substitutions have not yet emerged in the SHV family, providing insightful predictions about the potential future evolution of the enzyme. Detailed description of communities with secondary importance for protein function (community 3, 8, and 9) is provided in the supplementary information (Supplementary Fig. S6).

DyNoPy predicts mutation hotspots in SHV-1

DyNoPy detects critical mutation sites (L162 and N136) that are known to extend the range of substrates in other class A β-lactamases but have not yet emerged as variants in the SHV family. These sites have not been modified in SHV family because of their plausible central role within the communities as they are mediating couplings with key functional residues essential for catalytic activity and structural stability, indicating their critical role in protein function and the potential lower mutation rate. These findings provide insightful predictions about the potential future evolution of the enzyme, as well as plausible explanations for why these mutations have not yet appeared.

L162, positioned at the start of the Ω-loop and adjacent to the crucial catalytic residue E166, is assigned as the core residue for community 1 (Fig. 3A). While it remains conserved in SHV family, variants of L162 have been isolated in other class A β-lactamase and are known to expand the enzyme catalytic spectrum. Single amino acid substitution at L162 can intensify antibiotic resistance in BEL-1 (50), a class A ESBL clinical variant, exhibiting robust resistance to ticarcillin and ceftazidime (51). BEL-2 diverges from BEL-1 by single amino acid substitution (L162F) which alters the kinetic properties of the enzyme significantly and increases its affinity towards expanded-spectrum cephalosporins (52). The relationship between L162 and protein catalytic functions can be explained using DyNoPy model, as there are couplings with catalytic important residues M69, K73, E166, and K234. Moreover, the BEL case has confirmed that L162F mutation significantly destabilizes the overall protein structure, highlighting the crucial role of L162 in maintaining protein stability (50). DyNoPy accurately identifies the centrality of L162 by reporting its connections with 28 backbone residues, including nine hydrophobic node residues critical for protein stability. Among these, five hydrophobic residues are part of the α2 node: V75, L76, G78, V80, and L81, highlighting the contribution of L162 to the stability of the α2 helix (35).

Community 1, 4, and 5 of SHV-1 β-Lactamase. All the residues are depicted as spheres on the protein structure. The core residue for each community is highlighted in red, while purple is used to emphasize the secondary core residue. Residues that interact with both cores are coloured in light yellow. Functional important residues are marked in cyan. Hydrophobic nodes are enclosed with cyan boxes. A. Community 1 of SHV-1, comprising 33 residues with L162 being the primary core residue. B. Community 4 of SHV-1, containing 12 residues and is centred by G156. G156 and A146 are two functional important residues distant from the active site. G156 is 21.3L away from the catalytic S70. A146 is 16.8L away from S70. C. Community 5 of SHV-1, embracing 48 residues and showing a strong correlation between V103 and S106.

Just like L162, N136 undergoes advantageous mutations in other class A β-lactamases while remains highly conserved within the SHV family. It is the core residue for community 7 (Fig. 4B). This residue forms a hydrogen bond with E166, stabilizing the Ω-loop (53). Although DyNoPy did not detect this direct interaction between N136 and E166, the established relationship between N136 and N170 highlights the role of N136 in influencing E166. N170, an essential catalytic residue located on the Ω-loop, contributes to priming the water molecule for the deacylation step with E166 (54) and is directly coupled with N136. Due to the essential contribution of N136 in facilitating E166 to maintain its proper orientation, it was previously thought to be intolerant to mutations as substitution of Asparagine to Alanine at this position would make the enzyme lose its function completely (55). However, N136D substitution has emerged as a new clinical variant very recently in PenL, a class A β-lactamase, by increasing its ability in hydrolysing ceftazidime (55), suggesting that this site has potential to mutate. This gain of function is mainly triggered by the increased flexibility of the Ω-loop (55). DyNoPy correctly detect a dynamical relationship between N136 and the Ω-loop (residues 164-179). Six residues present in the Ω-loop participate within this community, including R164 and D179. These two residues are critical as they are forming the ‘bottleneck’ of the Ω-loop which is essential for the correct position of E166 (56). D179 is also a critical mutation site for SHV-1. Single amino acid substitutions like D179A, D179N, and D179G are enough for the extended spectrum phenotype (46).

Community 6 and 7 of SHV-1 β-Lactamase. All the residues are depicted as spheres on the protein structure. The core residue for each community is highlighted in red, while purple is used to emphasize the secondary core residue. Residues that interact with both cores are coloured in light yellow. Functional important residues are marked in cyan. A. Community 6 of SHV-1, comprising 30 residues with Y105 being the primary core residue. R205 is a functional important residue that is 20.6L away from the active site S70. B. Community 7 of SHV-1, containing 34 residues and is centred by N136.

DyNoPy detects residue couplings essential for protein stability

DyNoPy identifies residue couplings critical for protein functional motions, particularly associated with protein stability. These residue pairs exhibit strong relationships as they are not only directly coupled with each other, but also forms various indirect couplings via other residues. As a result, both residues are considered as core residues inside these communities. It is expected that disruption of these couplings through mutation could compromise collective motions essential for enzyme activity.

As the secondary core residues in community 1 (Fig. 3A) F72 is showing a strong coupling with the primary core residue L162 and also forms nine indirect couplings with L162, including via the catalytic K234. This network of direct and indirect relationships reveals the importance of F72 and L162 coupling in maintaining protein functional motions. Interestingly, previous studies identified a small hydrophobic cavity formed by L162 and F72, together with L139, and L148, which is essential for the stability of the active site (50). Notably, DyNoPy successfully recovers the key residues of this local hydrophobic cavity (L162, F72, and L148).

The strong interplay between V103 and S106, which are both residues on the α3-α4 loop, is seen in community 5 (Fig. 3C). These residues not only interact with each other directly but are also indirectly coupled via 22 other residues. This community emphasizes the significance of hydrophobic nodes in SHV stability and dynamics. Within the analysed 48 residues, 27 are hydrophobic, out of which 15 residues act as nodes critical for enzyme stabilization. Hydrophobic nodes stabilize their own secondary structures and interconnect to stabilize the overall protein (57). V103 and S106 themselves are hydrophobic nodes, stabilizing α3 helix and α4 helix respectively, and are strongly coupled with each other. In CTX-M, another class A enzyme, N106S is a common substitution that results in improved thermodynamic stability and compensate for the loss in stability of the variants (58). Interestingly, this residue is already a Serine in SHV, but still implies its pivotal role in protein stability.

DyNoPy provides valid explanations for mutation sites

During the evolution of β-lactamases, single mutations on specific sites that are distant from the functional sites have been observed to significantly alter protein catalytic functions. Additionally, single mutations on some surface exposed residues can dramatically increase protein stability. Understanding how these distant mutations impact function and stability becomes a major challenge in understanding protein evolutionary pathways. Communities extracted by DyNoPy show these residues linked with functional important residues, providing a rational for these mutation sites with unknown functions.

Mutations of G156 are limited but they lead to ESBL phenotype in the SHV family (46). G156 is the central residue for community 4 (Fig. 3B), but it is distant from the active site, over 20L away from the catalytic serine S70. Clinical variant SHV-27, has extended resistance ability towards cefotaxime, ceftazidime, and aztreonam (59). It differs from SHV-1 by single amino acid substitution G156D, suggesting that it has directly evolved from SHV-1 (59). Limited research has been done on position G156, and the understanding of how it affects the enzyme catalytic properties given that it is far away from the active site is still unclear. Based on our results, we suggest that this residue is essential for the overall protein function because of its 11 coevolved dynamic couplings with protein dynamics, including A146, another ESBL substitution site.

SHV-38, another ESBL that is capable of hydrolysing carbapenems, harbours a single A146V substitution compared to SHV-1 (60). Like G156, A146 is 16.8 L away from S70 but shows an ability in altering protein catalytic function. The A146-G156 residue pair shows a strong coevolutionary signal and strong correlation with protein overall dynamics, implying that there may compensatory mutations at these sites with potential to emerge in the SHV family in the future. These two residues are not connected to any catalytic residues but their coupling to functional dynamics can offer plausible explanation to ESBL activity of these two mutations.

Unlike other substitution sites that are adjacent to the active site, R205 is situated more than 20 L away from catalytic serine S70. Its side chain points outward from the protein, exposing to the solvent. The R205L substitution often co-occurs with other ESBL mutations and is thought to indirectly contribute to the ESBL phenotype by compensating for stability loss induced by other mutations (61). SHV-3 is an ESBL that exhibits significant resistance to cefotaxime and ceftriaxone (62). Two substitutions in this enzyme, R205L and G238S, extend its resistance profile (62). Thus, it is promising to see that DyNoPy detected these two mutation sites together within community 6 (Fig. 4A).

Y105 and R266 are the core residues for community 6. Y105 is situated on the α3-α4 loop positioned at the left side of the binding pocket. It is an important catalytic residue that recognizes and binds to the thiazolidine ring of penicillins or β-lactamase inhibitors (63). There is very limited information on the role of R266, except that it may stabilize the Ω-loop in the SHV family similar to the analogous T266 in TEM (64). G238 is coupled with an essential catalytic residue Y105, which further links with other catalytic functional residues: S70 and A237, and R266, a residue that known to stabilize the Ω-loop. This indicates that mutations on G238 would result in an alteration on protein catalytic function, as well as an increased flexibility of the protein, which strongly aligns with previous finding (62). Its linked mutation site R205 does not showing direct coupling with any catalytic residues. Instead, it is directly coupled with R266, which we mentioned as an Ω-loop stabilizer. Thus, it is not surprising that R205 substitution alone is never observed in nature (65), as it would not give significant evolutionary advantage to the protein.

Insights into unexplained functional sites of PDC-3

Unlike the extensively studied SHV-1, the functional roles of individual amino acids in PDC-3 remains largely unexplored. This gap in understanding serves as welcome challenge for interpreting the effects of mutations and the dynamic behaviour of PDC-3 from our results. Although several mutation hotspots, such as those on the Ω-loop (48), have been identified, very little is known about the specific contributions of individual amino acids on the functionality of PDC-3.

In PDC-3, mutations have primarily been reported in the Ω-loop. They enhance its flexibility to accommodate the bulky side chains of antibiotics, while deletions are more common in the R2-loop (43). DyNoPy detected five communities in total (Supplementary Fig. S4B) with all the four predominant Ω-loop mutations appeared in these communities. Community 3, 4 and 5 are explained in the Supplementary information (Fig. S7). Furthermore, DyNoPy also detected several previously unexplored Ω-loop residues.

G214, a known mutation site in PDC-3, is the core residue in community 1. Another two essential mutation sites: E219 and Y221 also participate in this community, directly coupled with G214 (Fig. 5A). G214 also has direct couplings with four other Ω-loop residues: A195, A197, G212, and L216. Previous results have demonstrated that substitutions of Glycine to Alanine or Arginine at 214 significantly destabilizes the Ω-loop (36). The strong correlation between G214 and these Ω-loop residues emphasizes the significant contribution of G214 towards the stability of the Ω-loop, which corroborates with previous results (36). Moreover, substitutions such as G214A and G214R and mutations on E219 and Y221 do not affect R2 loop flexibility, resulting in the smaller active site volume among variants (36) because none of the residues from the R2 loop are detected in this community offering plausible explanation to previously unexplained phenomenon.

Community 1 and 2 of PDC-3 β-Lactamase. All the residues are depicted as spheres on the protein structure. The core residue for each community is highlighted in red. Functional important residues are marked in cyan. A. Community 1 of PDC-3, comprising 36 residues with G214 being the primary core residue. B. Community 2 of PDC-3, containing 74 residues and is centred by G204.

G204 is the core residue of community 2, coupled with 73 other residues, most of which are distant from the catalytic site, suggesting plausible crucial role in overall protein stability like L162 in SHV-1 (Fig. 5B). G204, a newly emerged mutation site in the PDC family (66), is located on the short β-sheet β5a within the Ω-loop, near the hinge region between β8 and β9 just above the active site. The only known variant of G204 is PDC-466, which was derived from PDC-462 (A89V, Q120K, V211A, N320S), with an addition of G204D (66). Coupling of G204 to several catalytically important residues, including K67, K315, and T316 can suggest that mutations at this site can negatively impact catalytic power. This offers a plausible explanation of seeing fewer variants at this site and mutations at this site could have impact on hydrolysing capabilities of PDC variants. This should be confirmed by further experimental studies of variants of G204. Unlike G214, E219 and Y221 mutations which do not influence the dynamics of the R2 loop, substitutions on V211, a member of Ω-loop, has impact on dynamics of R2 loop because of its indirect couplings, through G204 to R2-loop residues (36). Two less critical substitution sites, H188 and V329, were also observed in community 2.

Conclusions

DyNoPy offers two distinct advantages over existing computational tools (24, 26): a) information on residue-residue coevolution can be directly used to detect the components of protein dynamics that have been preserved during evolution b) dynamic descriptors extracted from the MD ensembles can be used to identify the function-specific conserved dynamic couplings. These couplings are then easily modelled as a graph and network analysis is used to extract epistatic communities and assign roles to residues based on their importance in the graph model. The choice of a relevant descriptor of functional dynamics has an impact on the ability to detect couplings that are involved in functional dynamics. In systems where there is limited role of dynamics in the function, the analysis done with DyNoPy is equivalent to conventional coevolution analysis, which can be consider one limitation of our method.

Here we demonstrated how the choice of relevant global and local descriptors returns a higher number of effective couplings (greater than 0), and in turn leads to interpretable graph models and communities. In other systems, when multiple descriptors can be used to quantify functional conformational change, it is expected that they will differently modulate the effect of coevolution coupling, which will be reflected in a different structure of the associated graph models. This suggests the use of DyNoPy to generate comparative models in proteins with multiple functions associated to distinct dynamical changes.

Mutations of L162 and N136 have not yet emerged in SHV-1, but they are detected by DyNoPy as core residues for communities. These residues are strongly coupled with other functional important residues, which play critical roles in protein stability and catalytic activity. The identification of these couplings shows high consistency with previous studies and highlights the importance of L162 and N136 in SHV-1 functional dynamics. Given their central role in these communities, mutations in L162 and N136 can significantly alter protein function, suggesting their potential for future evolutionary changes. However, their strong relationships with these critical functional residues also suggest that mutation at these sites would need to be balanced to maintain protein function, providing an explanation for why such mutations have not yet emerged in SHV-1 (67). The ability of DyNoPy in detecting functionally important mutation sites was demonstrated via well-characterized mutation sites including R205 and G238 from SHV-1. Moreover, DyNoPy shows predictive ability on less-studied mutation sites such as G156 and A146, by detecting critical residue couplings that coevolved with functional motions.

Based on the knowledge we have gained from analysis of SHV-1 functional protein dynamics we suggest that in PDC-3, mutations at G204 because of its significant conserved dynamic couplings can lead to new ESBL/IRBL clinical variants. We suggest that DyNoPy can be used as a predictive tool to identify potential functional residues within this enzyme and guide future mutagenesis studies.

In summary, by integrating hidden evolutionary information with direct dynamic interactions, DyNoPy provides a powerful framework for identifying and analysing functional sites in proteins. The tool not only identifies key residues involved in local and global interactions, but also improves our ability to predict silent residues with previously unknown roles for future experimental testing. Our application of DyNoPy to broad-spectrum β-lactamases ESBLs and IRBLs demonstrates its potential to address key medical challenges such as antibiotic resistance by providing valid predictions on protein evolution.

Methodology

DyNoPy generates a graph representation of the protein structure that captures the couplings between amino acid residues contributing to the functional dynamics of the protein. Residues are represented as graph nodes, and conserved dynamic couplings are recorded as edges. Edge weights quantify the strength of these couplings. The model is built on two assumptions: residue pairs should have i) coevolved and their ii) time-dependent interactions correlate with a functional conformational change.

Therefore, edge weights (Jij) for residue i and j are calculated as: where γij is the scaled coevolution score and ρij is the degree of correlation with the selected functional conformational change. α and β are weights assigned to γij and ρij that have a sum of one. The relative weight of the scaled coevolution score (α) is set to 0.5 in this study. When either of the assumptions listed above is not met, Jij is set to zero.

Scaled coevolution scores

The occurrence of residue-residue coevolution can be estimated and quantified using probabilistic models of correlated mutations from deep multiple sequence alignments (MSA). DyNoPy supports generation of the MSA using the HH-Suite package (68) and calculation of scaled coevolution score (γij) using CCMpred (69) as per the protocol described in Bibik et al. (70). For SHV-1 and PDC-3 hhblits returned 18,174 sequences (Neff: 11.082) and 27,892 sequences (Neff: 9.951). Sequences were detected from the UniRef30 (v2022_02) database (71). First a pairwise residue coevolution matrix (C) is calculated, then these raw scores (Cij) are divided by the matrix mean (Equation 2). All scores (Sij) smaller than 1 are set to zero, and the remaining values are normalised by the maximum value (Equation 3):

Correlation with functional motions

The contribution of a residue pair to a selected functional motion is estimated by how much the change in interaction energy between the two residues over time is correlated with a collective variable (CV) describing the functional motion: where εij(t) is the pairwise non-bonded interaction energy (see details in Supplementary Information) and d(t) is the time-dependent value of the CV. Examples of CV and a discussion on the choice of the most relevant CV is presented in the results section. Correlation values smaller than 0.5 are set to 0. In absence of detectable contributions to the functional dynamics of the system, the couplings extracted by DyNoPy will describe a pure evolutionary model, and the community detection method presented below will be equivalent to a direct decomposition of the residue coevolution network into units.

Graph representation and analysis of conserved dynamic couplings

All pairwise conserved dynamic couplings (Equation 1) are collected into a square matrix J. A graph is built from J, using python-igraph v0.11 library (72). Nodes represent residues, and edges are drawn between nodes with positive Jij. Edge weights are set to Jij. The relative importance of the residues in this model of protein dynamics is calculated as eigenvector centrality of the nodes (73). Residues involved in extensive correlated dynamics with other highly connected residues have higher eigenvector centrality (EVC) scores. Groups of residues contributing to important collective motions are detected by community analysis of the graph structure. The Girvan-Newman algorithm is used to extract the community structure (74). A meaningful community should contain at least three residues. Applying network analysis on the combined dynamics-coevolution matrix helps us extracting higher-order interactions beyond pairwise coupling and detecting critical residues, which show multiple interactions with each other. Moreover, indirect long-range relationships, which would be hard to identify from numerical data, could be detected through community clustering. Community-based analysis offers a more comprehensive understanding of residue relationships and enables the visualization of residue couplings on the protein structure.

Adaptive Sampling Molecular Dynamics Simulations

MD simulation data was sourced from our previous studies (35, 36). To summarise, SHV-1 structural coordinates (PDB ID: 3N4I) were obtained from the Protein Data Bank and modified to the wild type by introducing the E104D mutation. Similarly, the PDC-3 structure was derived from PDC-1 (PDB ID: 4HEF) by a T105A substitution. Both enzymes were protonated at pH 7.0 using PropKa from the PlayMolecule platform (75). One disulfide bond between C77 and C123 was specified in SHV-1. Both structures were solvated with TIP3P water molecules in a periodic box with a box size of 10 Å. Ions were added to neutralize the overall charge of each system at 150mM KCl. Amber force field ff14SB was used for all MD simulations (76). After an initial minimisation of 1000 steps, both the enzymes were equilibrated for 5 ns in the NPT ensemble at 1 atmospheric pressure using the Berendsen barostat (77). The initial velocities for each simulation were sampled from the Boltzmann distribution at 300 K. Multiple Markov State Model (MSM)-based adaptively sampled simulations were performed for both proteins based on the ACEMD engine (78, 79). A canonical (NVT) ensemble with a Langevin thermostat (80) (damping coefficient of 0.1 ps−1) and a hydrogen mass repartitioning scheme were employed to achieve time steps of 4 fs. For SHV-1, each trajectory spanned 60 ns with a time step of 0.1 ns, with a total of 593 trajectories. In the case of PDC-3, 100 trajectories were collected, each containing 3000 frames, lasting 300 ns. To manage the extensive datasets efficiently, trajectories were strategically stridden to ensure that a minimum of 30,000 frames were preserved for each system. The resulting trajectories are summarized in Supplementary Table S4.

Calculation and Selection of Collective Variables

DyNoPy works on the assumption that time-dependent interactions between critical residues, either having significant structural change or not will correlate with functional conformational motions. Since MD simulation data is high-dimensional, a time-dependent collected variable (CV) is required to extract the most relevant information for the process under study. The usefulness of DyNoPy is dependent on the choice of the CVs. To guide the selection of CVs, we selected 12 distinct features: radius of gyration (Rg), the first principal component (PC1), partial PC1 (PC1_partial), the first time-lagged independent component (TC1), partial TC1 (TC1_partial), global root mean square deviation (gRMSD), partial RMSD (pRMSD), dynamical RMSD (dRMSD), global solvent accessible surface area (gSASA), partial SASA (pSASA), active site pocket volume, and the number of hydrogen bonds (hbond). A description of the CVs, including the calculation methods and the residues used to calculate the partial variables, is detailed in the Supplementary information. CVs were subsequently used as input features for DyNoPy. A good collective variable (CV) should appropriately describe protein functional motions. Thus, a CV that detects the highest number of residue couplings is expected to be the most suitable descriptor. The length of the MD simulations should be appropriate to effectively sample the desired functional process as described by the selected CV.